
\renewcommand{\abstractname}{Acknowledgements}
\begin{abstract}
\begin{doublespace}
First and undeniably foremost, I extend my heartfelt gratitude to my family,
whose encouragement and words of wisdom and guidance were arguably the most
important contributor to my success in graduate school. Mom and Dad, you also
sparked my scientific curiosity from a young age, and introduced me to so
many scientific and intellectual opportunities. Elyssa, your encouragement
and humor were always a much-needed reminder that light actually existed at
the end of the tunnel I was passing through. I'm so proud of you, and I'm glad
I have you as my sister.
\\\\
Of course, no mention of family would be complete without my colleagues and
mentors I had the privilege of working with on a daily basis. I'm happy the
pressure cooker of graduate school brought us all closer than any ordinary
coworkers. Dr. Martha Morton, your honest, uncensored insights on academic
research, analytical chemical instrumentation and human interaction were
always a pleasure to receive and will remain an invaluable aid in whatever
future I choose to pursue. To Matt Shortridge, Jaime Stark, Jenni Copeland,
Bo Zhang, Steve Halouska, Teklab Gebregiworgis, Darrell Marshall, Shulei Lei
and Jonathan Catazaro, I'm thankful for the opportunities I had to share
everything from scientific results to burdens of conscience with you. I
would not have chosen to endure the rigours of graduate school if it were
not for all of you. To the current Powers group, I challenge you to keep
our truly remarkable comradery alive as the next round of junior students
enters our clan. I have undoubtedly learned the most from all of you, and
I hope those who join us next will share in that truly sustaining
collaborative, friendly atmosphere that the Powers group thrives upon.
\\\\
Finally, I wish to sincerely thank my advisor, Dr. Robert Powers, and the
members of my supervisory committee, Drs. David Hage, Gerard Harbison,
Eric Dodds, and Stephen Scott for providing me with an environment where
I am free to pursue my intellectual curiosities, no matter how apparently
unrelated they were at the time to this dissertation. Thank you all for
your patient instruction and guidance during my time at the University of
Nebraska.
\end{doublespace}
\end{abstract}

